---
title: Dashboard
format: dashboard
---

# This code takes a long time to load FYI
```{r setup, include=FALSE}
library(tidyverse)
library(sf)
sv_votes <- read_csv("/Users/estebanlopez/Downloads/state_g24_sov_data_by_g24_srprec 2.csv")
sr_shp <- st_read("/Users/estebanlopez/Downloads/srprec_state_g24_v01_shp/srprec_state_g24_v01_shp.shp")
ab604_shp <- st_read("/Users/estebanlopez/Downloads/AB604/AB604.shp")


ab604_est_votes <- read_csv("data/derived/ab604_estimated_votes_area_weighted.csv")

dir.create("data/derived", recursive = TRUE)


print("Processing 2024 Map... (This will be slow)")


district_votes_2024 <- sv_votes %>%
  group_by(CDDIST) %>%
  summarise(
    DEMVOTE = sum(as.numeric(CNGDEM01), na.rm = TRUE) + 
              sum(as.numeric(CNGDEM02), na.rm = TRUE),
    REPVOTE = sum(as.numeric(CNGREP01), na.rm = TRUE) + 
              sum(as.numeric(CNGREP02), na.rm = TRUE)
  )

precinct_to_district_lookup <- sv_votes %>%
  select(SRPREC_KEY, CDDIST) %>%
  distinct() 
map_2024_geom <- sr_shp %>%
  st_transform(3310) %>%
  st_buffer(0) %>% 
  left_join(precinct_to_district_lookup, by = "SRPREC_KEY") %>%
  group_by(CDDIST) %>%
  summarise(geometry = st_union(geometry)) %>% 
  st_make_valid()
map_2024 <- map_2024_geom %>%
  left_join(district_votes_2024, by = "CDDIST") %>%
  mutate(share_A = DEMVOTE / (DEMVOTE + REPVOTE))
print("Processing AB 604 Map...")

map_ab604 <- ab604_shp %>%
  st_transform(3310) %>%
  st_make_valid() %>%
  left_join(ab604_est_votes, by = c("DISTRICT" = "ab604_district")) %>%
  mutate(share_A = DEMVOTE / (DEMVOTE + REPVOTE))
print("Saving processed files...")
saveRDS(map_2024, "data/derived/map_2024.rds")
saveRDS(map_ab604, "data/derived/map_ab604.rds")
saveRDS(district_votes_2024, "data/derived/district_votes_2024.rds")
saveRDS(ab604_est_votes, "data/derived/ab604_est_votes.rds")

print("Processing Complete!")
```


```{r setup, include=FALSE}
library(tidyverse)
library(sf)
library(ggpolypath)
library(knitr)   
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE)
DEM_REP_SCALE <- scale_fill_gradient2(
  name = "Dem Vote Share",
  low = "red", mid = "white", high = "blue",
  midpoint = 0.5,
  labels = scales::percent)
THEME_MAP <- theme_void() +
  theme(legend.position = "bottom")

compute_metrics <- function(df, votesA = "DEMVOTE", votesB = "REPVOTE") {
df2 <- df %>%
mutate(votes_A = .data[[votesA]],
votes_B = .data[[votesB]],
total = votes_A + votes_B,
share_A = votes_A / total,
share_B = votes_B / total,
winner = if_else(votes_A > votes_B, "A", "B"),
wasted_A = if_else(winner == "A", votes_A - (total/2 + 0.5), votes_A),
wasted_B = if_else(winner == "B", votes_B - (total/2 + 0.5), votes_B))
mean_share <- mean(df2$share_A, na.rm = TRUE)
median_share <- median(df2$share_A, na.rm = TRUE)
mean_median <- mean_share - median_share
total_wasted_A <- sum(df2$wasted_A, na.rm = TRUE)
total_wasted_B <- sum(df2$wasted_B, na.rm = TRUE)
efficiency_gap <- (total_wasted_A - total_wasted_B) / sum(df2$total, na.rm = TRUE)
list(mean_median = mean_median, efficiency_gap = efficiency_gap, seats_A = sum(df2$winner == "A", na.rm = TRUE), seats_B = sum(df2$winner == "B", na.rm = TRUE))
}
map_2024 <- readRDS("data/derived/map_2024.rds")
map_ab604 <- readRDS("data/derived/map_ab604.rds")
district_votes_2024 <- readRDS("data/derived/district_votes_2024.rds")
ab604_est_votes <- readRDS("data/derived/ab604_est_votes.rds")
stats_2024 <- compute_metrics(district_votes_2024)
stats_ab604 <- compute_metrics(ab604_est_votes)
summary_table <- tibble(
  Metric = c("Democratic Seats Won", "Mean-Median Gap", "Efficiency Gap (for Dems)"),
  `2024 Map` = c(
    stats_2024$seats_A,
    scales::percent(stats_2024$mean_median, accuracy = 0.1),
    scales::percent(stats_2024$efficiency_gap, accuracy = 0.1)),
  `AB 604 Map (Estimated)` = c(
    stats_ab604$seats_A,
    scales::percent(stats_ab604$mean_median, accuracy = 0.1),
    scales::percent(stats_ab604$efficiency_gap, accuracy = 0.1)))
DEM_REP_SCALE <- scale_fill_gradient2(
  name = "Dem Vote Share",
  low = "red", mid = "white", high = "blue",
  midpoint = 0.5,
  labels = scales::percent)
THEME_MAP <- theme_void() + theme(legend.position = "bottom")
plot_2024 <- ggplot(map_2024) +
  geom_sf(aes(fill = share_A), color = "gray50", linewidth = 0.1) + # Added color/linewidth for borders
  DEM_REP_SCALE +
  THEME_MAP +
  labs(title = "2024 Congressional Map")

plot_ab604 <- ggplot(map_ab604) +
  geom_sf(aes(fill = share_A), color = "gray50", linewidth = 0.1) + # Added color/linewidth for borders
  DEM_REP_SCALE +
  THEME_MAP +
  labs(title = "Proposed AB 604 Map")
table_2024 <- tibble(
  Metric = c("Democratic Seats Won", "Mean-Median Gap", "Efficiency Gap (for Dems)"),
  Value = c(
    stats_2024$seats_A,
    scales::percent(stats_2024$mean_median, accuracy = 0.1),
    scales::percent(stats_2024$efficiency_gap, accuracy = 0.1)))
table_ab604 <- tibble(
  Metric = c("Democratic Seats Won", "Mean-Median Gap", "Efficiency Gap (for Dems)"),
  Value = c(
    stats_ab604$seats_A,
    scales::percent(stats_ab604$mean_median, accuracy = 0.1),
    scales::percent(stats_ab604$efficiency_gap, accuracy = 0.1)))
```

# 2024 Map

```{r}
#| title: "2024 Metrics"
kable(table_2024, align = "lr")
#| title: "2024 Congressional District Map"
plot_2024
```

# AB604 Map
```{r}
#| title: "AB 604 Metrics"
kable(table_ab604, align = "lr")
#| title: "Proposed AB 604 Map"
plot_ab604
```

# Methodology

## Data Sources

My data came the 2024 precinct-level vote results and the shapefiles were from the California Statewide Database. 

## Estimation Process

 I used an area-weighted interpolation, in which I made several steps:

1.  First, I joined the 2024 vote data to the 2024 precinct map.
2.  Second, I spatially intersected this map with the new AB 604 district map.
3.  Next, I calculated the area of each of sections I made, and then divided by the original precinct's total area
4. I then used that to allocate a part of the precieinct's 2024 congressional votes to that piece
5. Lastly, I summed all votes by their new id to get the estimated totals 

## Caveats

There is of course the caveat that is assumed when using the area-weighted intersection : votes are assumed to be uniformly distributed across districts. This oversimplifies the intricacies of a population and their votes. Therefore, it isn't safe to say this is a prediction; instead, this data should be interpretted as an estimate 
